The variability of Betelgeuse explained by surface convection

### 2024-04-16 main.tex
Add files via upload

Betelgeuse is a prototypical red supergiant (RSG) known to be a semi-regular variable. Several periods can be found in the 
literature usually clustered around the so-called long secondary period (LSP) of 2000 days and other more or less well defined 
periods around 400 and 200 days. Such periods are often linked to radial pulsation modes,   and have been used to model 
the characteristics of the resonant cavity, hence the size of the star and eventually its evolutionary stage \citep{saio_evolutionary_2023}.

At the end of 2019, Betelgeuse reached a historical minimum in its luminosity, called the great dimming \citep{guinan_fall_2020}. 
From interferometric data, we speculate that this event was caused by a cloud of dust close to the line of sight \citep{montarges_dusty_2021}. 
Interferometric images show a huge drop in luminosity in the southern hemisphere of Betelgeuse, leading to this dimming. In this event it was clear that a change in brightness was not due to any kind of pulsation but to a change on the 
brightness distribution over the stellar disk due to the presence of large quantities of dust. While this may be seen as 
a singular event, it brings up the question whether all changes in brightness cannot be due to similar changes in the brightness 
distribution over the disk, unrelated to any pulsation phenomenon.

Betelgeuse is known to present large convective cells with lifetimes in the order of one to two years and measurable changes 
in the span of one week. Bright convection cells near disk center will increase the integrated brightness of the star when 
compared with other situations where such cells are found near the edges. So even without calling for the formation of 
dust or other changes in opacity, one may argue that the simple evolution of convective patterns of the star may create a variability. Such 
variability could be expected to be random, but will present quasi-periodicities related to the typical time scales of the 
convective patterns.

Since the end of the Great Dimming event, Betelgeuse has continue its random variation of brightness. 
Interestingly, 
it has been shown by \cite{jadlovsky_analysis_2023} and \cite{dupree_great_2022} that the periodicity of Betelgeuse has changed after the dimming. Using the light curves from AAVSO, 
the authors showed that before the dimming, the dominant period of Betelgeuse was the 400 days period. However, after the dimming, this period has changed and 
is now shorter, around 200 days. Therefore, after the dimming, there was a change in the behaviour of the atmosphere of Betelgeuse: the time scales on which Betelgeuse evolves are now shorter. Our analysis is based on this change of periodicity. 
Faced with the dramatic change in stellar parameters and age that such change in a pulsation period would 
require, our interest in an alternative explanation of the variability
in terms  of convective patterns arose. In order to address this question we shall seek for such typical periods of 2000, 400 and 200 days in  
observational proxies related to the convective activity but unrelated to pulsations and which cannot be directly linked to brightness.

Linear polarization in the atomic lines of the spectra of Betelgeuse, discovered by \cite{auriere_discovery_2016}, has been interpreted as the joint action of 
two mechanisms. First the depolarization of the continuum by atoms which absorb linearly polarized light and re-emit unpolarized light, the 
continuum photons being polarized by Rayleigh scattering. This depolarization produces signals with an azimuth symmetry over the disk that 
would  cancel out the net linear polarization on a homogeneous disk. Thus it must be combined with an inhomogenous disk to produce the net linear polarization 
signal observed. This interpretation suggested the possibility of mapping those brightness inhomogeneities. This has been done by \cite{lopez_ariste_convective_2018} 
even producing 3-dimensional images of the atmosphere of Betelgeuse \citep{lopez_ariste_three-dimensional_2022} when taking advantage of the different height of formation of 
different lines in the spectrum of Betelgeuse. The images produced compare very well with contemporaneous images made with interferometric 
techniques \citep{montarges_close_2016} and show clear convective patterns, akin to solar granulation.

Linear polarization observed in the atomic lines is therefore a proxy of convection, unrelated to radial 
pulsations but linked to the brightness inhomogeneities due to the convective patterns in Betelgeuse. If one can find the aforementioned periods 
in linear polarization, we must conclude that these periods are related to the convective activity which originates those linear polarization signals
and unrelated to any radial pulsation phenomena. This is the purpose of the present work.

\section{Searching for periodicities in the polarization spectra of Betelgeuse}

Betelgeuse has been observed for the last 13 years with Narval and Neo-Narval at the Telescope Bernard Lyot (TBL). These two instruments measure the polarization 
over the visible spectra of Betelgeuese (390-1000 nm) with high spectral resolution (R=65000) and high polarimetric sensitivity. Despite 
this high sensitivity, the signal-to-noise ratios are not sufficient to measure the weak polarization signals in individual atomic lines of the 
spectrum. These amplitudes are today known to be of the order of $10^{-4}$ times the continuum intensity, and only exceptionally do they 
reach amplitudes of $10^{-3}$ the continuum. In these exceptional observations, enough photons can be accumulated per spectral bin to 
see the linear polarization signal above noise \citep{auriere_discovery_2016}. Most commonly, the amplitudes of linear polarization are below noise levels. In those 
occasions we have to add up the signals of thousands of lines to reduce the noise and increase the signal-to-noise ratios. This line addition 
is done through a technique called Least-Squares Deconvolution (LSD) \citep{donati_spectropolarimetric_1997} and assumes that the 
linear polarization signal is similar 
in all spectral lines up to scale factors in the sampling and amplitude. This technique has been successfully used in the past to measure 
magnetic field distributions over stellar surfaces and it is now used to produce images of the brightness distributions in the photosphere 
of Betelgeuse and other red supergiants as cited in the Introduction.

@@

Trying to find periods in an astrophysical context carries the hurdle of unevenly spaced observed data. To overcome this issue, 
we used the Lomb-Scargle periodogram \citep{lomb_least-squares_1976,scargle_studies_1982}. For a given set of data, it fits sinusoidal 
using least-squares at each frequency between the first and the last observation. The better the fit, the higher the power attributed to the frequency. 
We applied this technique to the  LSD profiles of Stokes I, U and Q of Betelgeuse observed at the TBL during 
that span of 13 years.

We first computed the Lomb-Scargle periodogram at each wavelength of the LSD profile, between -25 km/s and 80 km/s to span the entire 
signal present in the LSD profile. \cite{lopez_ariste_convective_2018} found the most blueshifted signal towards -20 km/s, 
and interpreted it as  the maximum velocity of the rising plasma, seen from the Earth. The most redshifted signal was found at 40 km/s. 
This velocity is interpreted
as the rest velocity of the star in our reference frame. Seldom, signals at velocities larger than 40 km/s can be found 
and can be interpreted as either dark and cold plasma falling back to Betelgeuse, 
or to rising plumes of plasma located in the hidden face of Betelgeuse,
and rising so high that they appear above the limb.

@@

Figures \ref{LS intensity},\ref{LS Q},\ref{LS U} and \ref{LS linear polarization} show the Lomb-Scargle periodogram in normalised power of the LSD profile of respectively 
the intensity, stokes Q, stokes U and the linear polarization. 
%The abscissa is the frequency (in $\mathrm{days^{-1}}$), while the ordinate is the heliocentric radial velocity in km/s. 
For each figure, the upper panel is the Lomb-Scargle periodogram with data taken before the great dimming of Betelgeuse, that is from 2013 to the beginning of 2019. 
The lower panel is the Lomb-Scargle periodogram for the whole dataset, before and after the dimming: from 2013 to 2023. The white dashed lines mark the different 
periods found in the literature: the LSP at 2000 days, the fundamental pressure mode at 400 days and its first overtone mode at 200 days.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle Intensity.png}
    \caption{Lomb-Scargle periodogram of the LSD profile of intensity. 
    The upper panel is the Lomb-Scargle periodogram for data before the great dimming while the lower panel is the periodogram from the whole dataset.
    The ordinate refers to the velocity in the heliocentric radial velocity in km/s.
    The three white dashed lines represent respectively (from left to right) the LSP at 2000 days, the 400 days period and its first overtone at 200 days.  }
    \label{LS intensity}
\end{figure}

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle Stokes Q.png}
    \caption{Same as figure \ref{LS intensity} for stokes Q. }
    \label{LS Q}
\end{figure}

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle Stokes U.png}
    \caption{Same as figure \ref{LS intensity} for stokes U.}
    \label{LS U}
\end{figure}

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle linear polarization.png}
    \caption{Same as figure \ref{LS intensity} for linear polarization: $\sqrt{\mathrm{Q^2}+\mathrm{U^2}}$.}
    \label{LS linear polarization}
\end{figure}

@@

%\subsection{Intensity}
Figure \ref{LS intensity} shows the Lomb-Scargle periodogram of the LSD of the intensity profile. Interestingly, we recover the LSP at 2000 days, 
but located at a HRV of 20 km/s rather than over the whole profile. This periodicity is still present after the  dimming, but  it is located at a HRV between 0 and 10 km/s. 
There is also a weaker signal at the fundamental mode, around 400 days before the dimming. But this signal completely disappears after the dimming. From these observations
we find  hard to attribute 
the observed Lomb-Scargle power to a global period of Betelgeuse.
%, however it might be a hint that this periodicity has changed after the dimming. 
Our results 
in the intensity profile are consistent with the one from \cite{mathias_evolution_2018} who used 2D Fourier analysis and found that the dominant period was the LSP.

\subsection{Stokes Q and U}

Figure \ref{LS Q} and \ref{LS U} show the Lomb-Scargle periodogram of both stokes Q and U. It is clear that the periodogram is completely different 
from the intensity one. A strong signal is present in fig. \ref{LS Q} around 200 days. Once again, this signal is weaker after the dimming.
Concerning stokes U in fig. \ref{LS U}, it appears that we only recover the LSP, which once again, is weaker after the great dimming. 
Those periodograms rise a lot of questions: why is the 200 days period only present in stokes Q? Why not in stokes U or intensity? 
We will discuss about those questions in the next section. It also appears that stokes Q is
much more sensitive to the 200 days period than the LSP. From this, we can conclude that stokes
Q evolves on a time scale much more closer to the first overtone 
mode than the LSP or even the fundamental mode. Since Stokes Q is sensitive to different periods, we must conclude that it is due to other mechanisms
than the one behind the periodicities observed in intensity. Linear polarization in Betelgeuse is interpreted to be a proxy of  convective cells,
so we can speculate that the signal present in the periodogram 
may be related to  the lifetime of convective cells. Such hypothesis is comforted by what we see in the 
LSD profiles of  Betelgeuse, whose shape is overhauled  in typical time scales of  a few months.
Such time scales observed in the Stokes profiles and attributed to the convective time scales 
appear to explain the strong signal present at 200 days in the Lomb-Scargle periodogram of Stokes Q.

\subsection{The linear polarization}

Figure \ref{LS linear polarization} shows the Lomb-Scargle periodogram of the total linear polarization of Betelgeuse: $\sqrt{\mathrm{Q^2+U^2}}$. 
Interestingly, the periodogram shows a strong power at 400 days and 2000 days. This signal is much more weaker after the dimming.
While the LSP is already present in fig. \ref{LS intensity}, here we recover the 400 days periodicity, often invoked in the literature. 
Nonetheless, the power of this period peaks at a HRV of around 40 km/s. We recall that this velocity is the rest velocity of the star. 
Every signal beyond this velocity is either due to  plasma sinking in Betelgeuse, or to plumes of plasma rising beyond and above the limb of the star. 
Such plumes from the back hemisphere have been often seen in other red supergiants like $\mu$ Cep \citep{lopez_ariste_height_2023}. 
This suggests that both signals at these period of 400 days found in intensity and total linear polarisation 
could well be due to the periodic appearance of plumes in the back hemisphere rising high enough to appear above
the visible limb and generating a supplementary source of brightness. Once again we find a plausible explanation 
of these periods unrelated to pulsations, but rather to the convective dynamics of the star.

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle Photo-center.png}
    \caption{Lomb-Scargle periodogram of the 100 photo-center displacement of Betelgeuse. Each blue curve corresponds to a Lomb-Scargle periodogram of one photo-center
     displacement. The black line is the average of the 100 periodogram. The black dashed lines represent the 2000, 400 and 200 days period respectively 
     (from left to right). In the upper panel, only data before the dimming were used to compute the Lomb-Scargle periodogram, while for the lower panel, the whole dataset was used. }
    \label{LS photocenter}
\end{figure}

Figure \ref{LS photocenter} shows the Lomb-Scargle periodogram of the photo-center displacement of each one of those 100 series (blue lines) and the average Lomb-Scargle periodogram (black line). 
As before, the upper panel shows the Lomb-Scargle periodogram from data before the great dimming only, whereas we used the whole set of data for the lower panel.
The black dashed lines mark the 2000, 400 and 
200 days periods respectively. First, we find that the Lomb-Scargle power is high around the LSP, though
uncomfortably close to the auto-correlation peak.
We also notice a decrease in the Lomb-Scargle power from  before to after the dimming, in 
concurrence to what was seen in the periodograms of the LSD profiles. Before it, there are two smalls bumps around the 400 days period. 
But those bumps almost completely disappear after the dimming. Even though it is 
hard to identify the 400 days period from those bumps alone,
we still find that there is a difference before and after the dimming. 
And this is concluded  from images that exclusively involve surface convection.
The Lomb-Scargle periodogram of the photo-center displacement points towards a variability explained by convection only,
and this is consistent with what we obtained from linear polarization.
%Both the photo-center displacement and the LS of the linear polarization point toward a variability explained by surface convection.

\subsection{The variability of the light curve of Betelgeuse}

After examining the periods found by the Lomb-Scargle technique on the polarization data, and its derivatives, obtained by the TBL over the last 
13 years, it is worth putting them in the context of the periods found traditionally on light curves over the same period of time. Two aspects 
are of our interest in this comparison: the behaviour of the light curve before and after the great dimming, and the amplitudes of the peaks 
at the main cited periods.

@@

We have recovered the light curve of Betelgeuse in the visible from the AAVSO database for the last 13 years (Figure \ref{light curve Betelgeuse}).
Before the great dimming, the magnitude of Betelgeuse evolves on a timescale of a year, whereas after the dimming, the variability of Betelgeuse has changed. 
It is clear from fig. \ref{light curve Betelgeuse} that the 
variability of Betelgeuse is evolving on a timescale shorter than one year. This qualitative change has been pointed out before.
In the literature, it is reported that before the great dimming the main period of Betelgeuse was of 400 days \citep{kiss_variability_2006}. 
This period is often associated to the fundamental pressure mode. After the dimming, however this period appears to have disappeared 
and only the period at 200 days,  associated to the first overtone of the fundamental mode, is visible since \citep{dupree_great_2022}.
No explanations based upon pulsations have been offered concerning this change of variability after or perhaps because of the dimming. But this 
is a behaviour that fully concurs with what we find in the polarization data, and which cannot be related to any stellar pulsation but rather 
convective dynamics. Beyond the actual values of the periods found in a Lomb-Scargle periodogram, we also see a correspondence between polarization 
and the light curve in these changes in the observed periods before and after the great dimming. Such concurrence points to an explanation 
in terms of convective dynamics also for the variability of the light curve.

It may be argued that the periodograms presented in the previous sections and, in particular, those  of the photo-center displacements, 
present low amplitude peaks at the referred main periods. One may wonder whether we are over-interpreting mere fluctuations of the periodograms, 
while the peaks seen in the literature from analysis of the light curve are unmistakable. It is therefore worth to take advantage of the 
availability of the AAVSO data to perform a Lomb-Scargle periodogram of the light curve in the same period covered by our polarization data and,
furthermore, taking into consideration data around the dates for which TBL data is available. Periodograms  with such constraints are 
shown in Fig.\ref{LSAAVSO}. Even when the full dataset for the period is taken into account (red line), the periods present small amplitudes, and are 
fully and favourably comparable with the periodograms presented above. This is even more dramatic when the periodogram is made from data at the exclusive 
dates at which TBL data is available (blue line). It stands out that the low amplitude peaks are rather due to both the sparse cadence and short span of 
time of the TBL data series rather than to a physical absence. Actually, the present tests puts more emphasis on the importance of the 
strong peaks seen at selected wavelengths in the LSD profiles, strong when compared to the peaks at similar periods arising from the light curve.

>>>>

Betelgeuse is a prototypical red supergiant (RSG) known to be a semi-regular variable. Several periods can be found in the 
literature usually clustered around the so-called long secondary period (LSP) of 2000 days, and other shorter periods of 400 and 200 days. The origin of the LSP remains poorly known, some tried to explain the LSP by the lifetime of the large convective cells present at the surface of RSG (e.g \cite{stothers_giant_2010}) or by the magnetic field \citep{wood_long_2004}. The shorter periods of 400 and 200 days seem to have a different origin than the LSP. By studying a large sample of RSG, \cite{kiss_variability_2006} attributed the 400 days period to radial pulsation modes, the 200 days period being attributed to the first overtone \citep{joyce_standing_2020}, which was in agreement with the historical work of \cite{stothers_pulsation_1969}.\\

Besides radial pulsation, surface activity leading to random brightness variation was also mentioned to explain the variability of Betelgeuse, being firstly theorised by \cite{schwarzschild_scale_1975}. Betelgeuse is known to present large convective cells with lifetimes in the order of one to two years \citep{lopez_ariste_convective_2018} and measurable changes 
in the span of one week. Bright convection cells near disk center will increase the integrated brightness of the star when 
compared with other situations where such cells are found near the edges. So even without calling for the formation of 
dust or other changes in opacity, one may argue that the simple evolution of convective patterns of the star may create a variability. Such 
variability could be expected to be random, but will present quasi-periodicities related to the typical time scales of the 
convective patterns. \\

%more or less well defined 
%periods around 400 and 200 days. 
%Such periods are often linked to radial pulsation modes,   and have been used to model 
%the characteristics of the resonant cavity, hence the size of the star and eventually its evolutionary stage \citep{saio_evolutionary_2023}.

At the end of 2019, Betelgeuse reached a historical minimum in its luminosity, called the great dimming \citep{guinan_fall_2020}. 
From interferometric data, it is speculated that this event was caused by a cloud of dust close to the line of sight \citep{montarges_dusty_2021}. 
Interferometric images show a drop in luminosity in the southern hemisphere of Betelgeuse, that might be caused by a mass loss event and leading to this dimming \citep{dupree_great_2022}. Other theories such as a drop in temperature was also mentioned by \cite{harper_photospheric_2020}, or an increase in the molecular opacity \citep{kravchenko_atmosphere_2021}. In this event it was clear that a change in brightness was not due to any kind of pulsation but to a change on the 
brightness distribution over the stellar disk.
%due to the presence of large quantities of dust.
Since the end of the Great Dimming event, Betelgeuse has continue its random variation of brightness. 
Interestingly, 
it has been shown by \cite{jadlovsky_analysis_2023} and \cite{dupree_great_2022} that the periodicity of Betelgeuse has changed after the dimming. Using the light curves from AAVSO, 
the authors showed that before the dimming, the dominant period of Betelgeuse was the 400 days period, while after the great dimming the main periods of Betelgeuse have shortened, oscillating between 97 days and 230 days \citep{dupree_great_2022}, revealing a change in the behaviour of the atmosphere of Betelgeuse.
While the great dimming may be seen as 
a singular event, it brings up the question whether all changes in brightness cannot be due to similar changes in the brightness 
distribution over the disk, unrelated to any pulsation phenomenon. 
Our interest in an alternative explanation of the variability
in terms  of convective patterns arose. In order to address this question we shall seek for such typical periods of 400 and 200 days in  
observational proxies related to the convective activity.\\
%but unrelated to pulsations and which cannot be directly linked to brightness.
%Faced with the dramatic change in stellar parameters and age that such change in a pulsation period would 
%require,

Linear polarization in the atomic lines of the spectra of Betelgeuse, discovered by \cite{auriere_discovery_2016}, has been interpreted as the joint action of 
two mechanisms. First the depolarization of the continuum by atoms which absorb linearly polarized light and re-emit unpolarized light, the 
continuum photons being polarized by Rayleigh scattering. This depolarization produces signals with an azimuth symmetry over the disk that 
would  cancel out the net linear polarization on a homogeneous disk. Thus it must be combined with an inhomogenous disk to produce the net linear polarization 
signal observed. This interpretation suggested the possibility of mapping those brightness inhomogeneities. This has been done by \cite{lopez_ariste_convective_2018} 
even producing 3-dimensional images of the atmosphere of Betelgeuse \citep{lopez_ariste_three-dimensional_2022} when taking advantage of the different height of formation of 
different lines in the spectrum of Betelgeuse. The images produced compare very well with contemporaneous images made with interferometric 
techniques \citep{montarges_close_2016} and show clear convective patterns, akin to solar granulation.\

Linear polarization observed in the atomic lines is therefore a proxy of convection, unrelated to radial 
pulsations but linked to the brightness inhomogeneities due to the convective patterns in Betelgeuse. If one can find the aforementioned periods 
in linear polarization, we must conclude that these periods are related to the convective activity which originates those linear polarization signals.
%and unrelated to any radial pulsation phenomena. 
This is the purpose of the present work. In section \ref{Section 2} we describe the dataset of linear polarization obtained with the TBL and the Lomb-Scargle periodogram used to derive periods. In section \ref{section 3} we associate the 200 days period to convective timescale of the smallest granules and the 330 days periodicity to the largest granules, and we provide an explanation in the change of variability of Betelgeuse before and after the great dimming. In section \ref{Conclusion}, we summup our work.

\section{Searching for periodicities in the polarization spectra of Betelgeuse}
\label{Section 2}

Betelgeuse has been observed for the last 10 years with Narval and Neo-Narval at the Telescope Bernard Lyot (TBL). These two instruments measure the polarization 
over the visible spectra of Betelgeuese (390-1000 nm) with high spectral resolution (R=65000) and high polarimetric sensitivity. Despite 
this high sensitivity, the signal-to-noise ratios are not sufficient to measure the weak polarization signals in individual atomic lines of the 
spectrum. These amplitudes are today known to be of the order of $10^{-4}$ times the continuum intensity, and only exceptionally do they 
reach amplitudes of $10^{-3}$ the continuum. In these exceptional observations, enough photons can be accumulated per spectral bin to 
see the linear polarization signal above noise \citep{auriere_discovery_2016}. Most commonly, the amplitudes of linear polarization are below noise levels. In those 
occasions we have to add up the signals of thousands of lines to reduce the noise and increase the signal-to-noise ratios. This line addition 
is done through a technique called Least-Squares Deconvolution (LSD) \citep{donati_spectropolarimetric_1997} and assumes that the 
linear polarization signal is similar 
in all spectral lines up to scale factors in the sampling and amplitude. This technique has been successfully used in the past to measure 
magnetic field distributions over stellar surfaces and it is now used to produce images of the brightness distributions in the photosphere 
of Betelgeuse and other red supergiants as cited in the Introduction.

@@

\subsection{The variability present in the stokes parameters}

Trying to find periods in an astrophysical context carries the hurdle of unevenly spaced observed data. To overcome this issue, 
we used the Lomb-Scargle periodogram \citep{lomb_least-squares_1976,scargle_studies_1982}. For a given set of data, it fits sinusoidal 
using least-squares at each frequency between the first and the last observation. The better the fit, the higher the power attributed to the frequency. 
We applied this technique to the  LSD profiles of Stokes $I$, $U$, $Q$ and to the total linear polarization of Betelgeuse observed by the TBL during 
that span of 10 years.

We first computed the Lomb-Scargle periodogram at each wavelength of the LSD profile, between -25 km/s and +60 km/s to span the entire 
signal present in the LSD profile. \cite{lopez_ariste_convective_2018} found the most blueshifted signal towards -20 km/s, 
and interpreted it as  the maximum velocity of the rising plasma, seen from the Earth. The most redshifted signal was found at +40 km/s. 
This velocity is interpreted
as the rest velocity of the star in our reference frame. Seldom, signals at velocities larger than +40 km/s can be found 
and can be interpreted as either dark and cold plasma falling back to Betelgeuse, 
or to rising plumes of plasma located in the hidden face of Betelgeuse,
and rising so high that they appear above the limb.

@@

Figures \ref{LS intensity},\ref{LS Q},\ref{LS U} and \ref{LS linear polarization} show the Lomb-Scargle periodogram of the LSD profile of respectively 
the intensity, stokes $Q$, stokes $U$ and the linear polarization. 
%The abscissa is the frequency (in $\mathrm{days^{-1}}$), while the ordinate is the heliocentric radial velocity in km/s. 
For each figure, the uper panel is the Lomb-Scargle periodogram of each velocity bin (black line) and the total average (red line). The green line corresponds to the window function. The lower panel is the Lomb-Scargle periodogram represented for each velocity bin, while the white dotted line refer to 400 and 200 days. The right panel is the mean profile of each stokes parameter for each velocity bin.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle Intensity.pdf}
    \caption{Lomb-Scargle periodogram of the LSD profile of intensity.
    The upper panel is the Lomb-Scargle periodogram for each velocity bin (black line) and the average (red line). The green line is the window function.
    In the lower pannel, the two white dashed lines represent respectively the 400 days and the 200 days periods. The right panel is the average intensity profile for each velocity bin.}
    \label{LS intensity}
\end{figure}

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle Stokes Q.pdf}
    \caption{Same as figure \ref{LS intensity} for stokes Q. }
    \label{LS Q}
\end{figure}

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle Stokes U.pdf}
    \caption{Same as figure \ref{LS intensity} for stokes U.}
    \label{LS U}
\end{figure}

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle linear polarization.pdf}
    \caption{Same as figure \ref{LS intensity} for linear polarization: $\sqrt{\mathrm{Q^2}+\mathrm{U^2}}$.}
    \label{LS linear polarization}
\end{figure}

@@

%\subsection{Intensity}
Figure \ref{LS intensity} shows the Lomb-Scargle periodogram of the LSD of the intensity profile. The 400 and 200 days period seem to be recovered by the periodogram. However, the 400 days period is uncomfortably close to the peak of the window function, while the 200 days period is not. We also notice that the two periods are present at the same HRV. Interestingly, \cite{mathias_evolution_2018} also found this 200 days periodicity with a shorter observation period.

Figure \ref{LS Q} and \ref{LS U} show the Lomb-Scargle periodogram of both stokes $Q$ and $U$. It is clear that the periodogram is completely different 
from the intensity one. A strong signal is present in fig. \ref{LS Q} around 200 days. Concerning stokes U in fig. \ref{LS U}, it appears that the main period is around $0.003 \ \mathrm{days^{-1}}$ (330 days). Other period such as the one at 200 days or 250 days ($0.004 \ \mathrm{days^{-1}}$) are present but difficult to trust. From stokes $Q$ and $U$, we recover the 200 days period and the 330 days period is more or less close to the 400 days period found in the literature. 
Figure \ref{LS linear polarization} shows the Lomb-Scargle periodogram of the total linear polarization of Betelgeuse: $\sqrt{\mathrm{Q^2+U^2}}$. 
Interestingly, the periodogram shows a strong power at both 330 days and 200 days, which are the period that we have found in both stokes $Q$ and $U$.

%Nonetheless, the power of this period peaks at a HRV of around 40 km/s. We recall that this velocity is the rest velocity of the star. 
%Every signal beyond this velocity is either due to  plasma sinking in Betelgeuse, or to plumes of plasma rising beyond and above the limb of the star. 
%Such plumes from the back hemisphere have been often seen in other red supergiants like $\mu$ Cep \citep{lopez_ariste_height_2023}. 
%This suggests that both signals at these period of 400 days found in intensity and total linear polarisation 
%could well be due to the periodic appearance of plumes in the back hemisphere rising high enough to appear above
%the visible limb and generating a supplementary source of brightness. Once again we find a plausible explanation 
%of these periods unrelated to pulsations, but rather to the convective dynamics of the star.

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle Photo-center.pdf}
    \caption{Lomb-Scargle periodogram of the 100 photo-center displacement of Betelgeuse. Each blue lines correspond to a Lomb-Scargle periodogram of one photo-center
     displacement. The black line is the average of the 100 periodogram. The red dashed lines represent the 400 and 200 days period respectively 
     (from left to right). As before, the green line is the window function.}
    \label{LS photocenter}
\end{figure}

Figure \ref{LS photocenter} shows the Lomb-Scargle periodogram of the photo-center displacement of each one of those 100 series (blue lines) and the average Lomb-Scargle periodogram (black line), 
%As before, the upper panel shows the Lomb-Scargle periodogram from data before the great dimming only, whereas we used the whole set of data for the lower panel.
while the black dashed lines mark the 400 and 
200 days periods respectively. As before, the window function is represented by the green line. Interestingly, we find two bumps around the 400 days period. We cannot directly confirm that we recover the 400 days period from those small bumps, but since polarimetry imaging involves only surface convection, this is an argument that points toward a variability explained by surface convection. 
%First, we find that the Lomb-Scargle power is high around the LSP, though
%uncomfortably close to the auto-correlation peak.
%We also notice a decrease in the Lomb-Scargle power from  before to after the dimming, in 
%concurrence to what was seen in the periodograms of the LSD profiles. Before it, there are two smalls bumps around the 400 days period. 
%But those bumps almost completely disappear after the dimming. Even though it is 
%hard to identify the 400 days period from those bumps alone,
%we still find that there is a difference before and after the dimming. 
%And this is concluded  from images that exclusively involve surface convection.
%The Lomb-Scargle periodogram of the photo-center displacement points towards a variability explained by convection only,
%and this is consistent with what we obtained from linear polarization.
%Both the photo-center displacement and the LS of the linear polarization point toward a variability explained by surface convection.

\section{The variability of the light curve of Betelgeuse}
\label{section 3}
%Before summing up the conclusions of this work, we can afford proposing alternative scenarios for the observed periodicities in the light curve, 
%the LSD profiles and the displacement of the photo-center of the inferred photospheric image reconstructions. 
The presence of the same 
periods and, after taking into consideration the time span and sparsity of the TBL data series, the larger amplitude of the peaks in
polarization data  than in the light curve point to a common origin, one that is rather based upon convective dynamics. Since we associate stokes $Q$ and $U$ to proxies of convection, we associate the 200 days period to the typical timescales on which the Stokes parameters evolve. 
The LSD profiles of stokes $Q$ or $U$ will slightly change when we observe Betelgeuse every month, but are completely overhauled 
on time scales of several months. The Lomb-Scargle periodograms appear to capture this spectral dynamic around periods of 200 days.
 Such changes in the profiles are directly linked to the dynamics of the surface,  to  the evolution and movements 
of convective cells over the visible hemisphere. So the 200-day period appears to be related to the evolution of the convective patterns 
on the surface of Betelgeuse.  The fact that this peak is strongly present in stokes $Q$ but less in $U$ points toward a more stochastic origin, which can fully be explained by convection.\

The other period found in the linear polarization and stokes $U$, the 330 days periods, is more or less close to the 400 days period often present in the literature (e.g \cite{kiss_variability_2006} found a period $388 \pm 30$ days). It has been shown by \cite{lopez_ariste_convective_2018} that large convective cells can live up to one year, thus this period would be associated to the timescale on which the largest granules evolve. This typical timescale has also been observed in the numerical simulations and might play an important role in the determination of stellar distances through the displacement of the photo-center \citep{chiavassa_probing_2022}. Once again, the 330 days periodicity being more present in stokes $U$ than $Q$ unravels a stochastic process that is due to convective motions. \

Finally, the intensity profile seems to match the 400 days periods, but is uncomfortably close to a peak of the window function of the TBL and a secondary peak close to 200 days. Since we are able to recover the 200 days period in the stokes paramters, we attribute this period to the convective timescales, but concerning the 400 days period, it could be a mix of the typical convective timescales of large granules associated with stochastic pulsations exited by convective cavities. Up to this point, the model developed to interpret linear polarization does not involve pulsations, that is why we interpret the 200 days period as the convective turnover timescale. The 400 days period involve the intensity profile, thus the 400 days period is probably caused by both convective timescales and stochastic pulsations.
 %As showed by
%\cite{lopez_ariste_convective_2018}, some of those convective cells can live up for years. We may speculate
%that the LSP may be related to those long living structures. However the present span of data from the TBL is insufficient to do more than
%just speculate. It should be noticed that  
% the peak is not present in stokes U, and this may point to a more stochastic event than just long-living structures.

\subsection{The light curve of Betelgeuse from the point of view of convection}

After examining the periods found by the Lomb-Scargle technique on the polarization data, and its derivatives, obtained by the TBL over the last 
10 years, it is worth putting them in the context of the periods found traditionally on light curves over the same period of time. Two aspects 
are of our interest in this comparison: the behaviour of the light curve before and after the great dimming.
%and the amplitudes of the peaks 
%at the main cited periods.

@@

We have recovered the light curve of Betelgeuse in the visible from the AAVSO database for the last 10 years (Figure \ref{light curve Betelgeuse}).
Before the great dimming, the magnitude of Betelgeuse evolves on a timescale of a year, whereas after the dimming, the variability of Betelgeuse has shortened. 
It is clear from fig. \ref{light curve Betelgeuse} that the 
variability of Betelgeuse is evolving on a timescale shorter than one year. This qualitative change has been pointed out before,
before the great dimming the main period of Betelgeuse was of 400 days \citep{kiss_variability_2006}. 
This period is often associated to the fundamental pressure mode. After the dimming however, this period appears to have disappeared 
and only timescales shorter than 230 days are visible since \citep{dupree_great_2022}. 
No explanations have been offered concerning this change of variability after or perhaps because of the dimming, thus we can afford proposing alternative scenarios for the observed periodicities in the light curve. \

RSGs encounter mass loss event, where due to stellar winds, plasma is ejected in the interstellar medium. Those events have already been seen in RSG like $\mu$Cep, where \cite{lopez_ariste_height_2023} found an excess of linear polarization beyond the limb of the star and associated it to rising plumes of plasma in the back hemisphere of the star. In the case of Betelgeuse, this kind of event happened only once or twice during the last 10 years, with no change in the variability of the star. However, since the great dimming, the variability of Betelgeuse has changed, meaning that this event was such different from the usual mass loss event, that it hustled the dynamics of the photosphere. \

Considering a scenario where the variability of Betelgeuse is mostly governed by convection, before the great dimming, the variability of the star was linked to the typical timescale on which convection occurs, that is 200 days for the small structures and roughly 350 days for the large structures. The great dimming impacted the behaviour of the photosphere, such that the largest structures do not evolve on timescale of the order of 350 days but rather 200 days, that is the smallest convective structures. We hypothesise that since the dimming, the photosphere is not in equilibrium and the largest convective structures are destroyed by the turbulent motion of the photosphere. Therefore, it is hard to identify periods after the great dimming, which \cite{dupree_great_2022} noticed. We expect to recover the 400 days variability within the next years, which corresponds to the time for the photosphere to go back to equilibrium.  
%But this 
%is a behaviour that fully concurs with what we find in the polarization data, and which cannot be related to any stellar pulsation but rather 
%convective dynamics. Beyond the actual values of the periods found in a Lomb-Scargle periodogram, we also see a correspondence between polarization 
%and the light curve in these changes in the observed periods before and after the great dimming. Such concurrence points to an explanation 
%in terms of convective dynamics also for the variability of the light curve.
